---
output:
  BiocStyle::html_document
---

<!---
Because we split the analysis pipeline in different independent files,
to speed up processing it, here in the setup block we load libraries and
objects that were loaded or produced in the previously processed file,
and which are necessary in this file.
--->

```{r setup, cache=FALSE, echo=FALSE, results='hide', message=FALSE}
library(knitr)
library(SummarizedExperiment)
library(edgeR)
library(geneplotter)

opts_chunk$set(cache=TRUE,
               cache.path="cache/file2",
               cache.extra=R.version.string,
               autodep=TRUE,
               fig.align="center",
               comment="")

thca <- readRDS("./seTHCA.rds")
dge <- readRDS(file.path("./", "dge.rds"))
thca.filt.unnorm <- readRDS(file.path("./", "thca.filt.unnorm.rds"))
dge.filt.unnorm <- readRDS(file.path("./", "dge.filt.unnorm.rds"))
thca.filt <- readRDS(file.path("./", "thca.filt4.rds"))
dge.filt <- readRDS(file.path("./", "dge.filt4.rds"))
```

# Differential expression

As we have seen in batch identification, samplevial cluster our dataset into two groups which can be confunding with the interest outcome. As we have not enough sample size for each samplevial to adjust the data, we are going to perform a Surrogate Variable Analysis (SVA) in order to try to capture sources of heterogeneity, such as non-biological variability by batch effects.

We perform a simple examination of expression changes and their associated p-values using the R/Bioconductor package [sva](http://bioconductor.org/packages/sva).

First, we observe how many differential expressed genes do we have without any batch adjustment

```{r}
library(sva)
mod <- model.matrix(~thca.filt$type, data = colData(thca.filt))
mod0 <- model.matrix(~1, data = colData(thca.filt))
pv <- f.pvalue(assays(thca.filt)$logCPM, mod, mod0)
sum(p.adjust(pv, method="fdr") < 0.01)
#thca.filt$pval <- p.adjust(pv, method="fdr") < 0.01
```

There are `r sum(p.adjust(pv, method="fdr") < 0.01)` genes changing significantly
their expression at FDR < 1%. In Figure \@ref(fig:pdist) below we show the distribution of the resulting p-values.
We use FDR because can accept some false positives and be more laxe. 

```{r pdist, echo=FALSE, out.width="400px", fig.cap="Distribution of raw p-values for an F-test on every gene between tumor and normal samples."}
hist(pv, main="Histogram before batch adjustment", las=1)
```

This histogram shows that most of the genes have a p-value less than 0.1, so they are considered as diferential expressed genes.

Now, we svestimate surrogate variables using the `sva()` function.

```{r}
assays(thca.filt)$logCPM <- cpm(dge.filt4, log=TRUE, prior.count=3)
sv <- sva(assays(thca.filt)$logCPM, mod, mod0)
sv$n
names(sv)
```

The SVA algorithm has found `r sv$n` surrogate variables. That way, we are going to use them to assess againt the extent of differential expression this time adjusting for these surrogate variables.

```{r}
modsv <- cbind(mod, sv$sv)
mod0sv <- cbind(mod0, sv$sv)
pvsv <- f.pvalue(assays(thca.filt)$logCPM, modsv, mod0sv)
sum(p.adjust(pvsv, method="fdr") < 0.01)
```

We have increased the number of changing genes to `r sum(p.adjust(pvsv, method="fdr") < 0.01)` with the batch adjustment for batch effects.
Figure \@ref(fig:psvdist) shows the resulting distribution of p-values.

```{r psvdist, echo=FALSE, out.width="400px", fig.cap="Distribution of raw p-values for an F-test on every gene between tumor and normal samples, adjusting for surrogate variables estimated with SVA."}
hist(pvsv, main="Histogram after batch adjustment", las=1)
```

In this histogram, we can observe there is more significant differential expressed genes with FDR less than 1%.

Once we have applied the batch adjustment, we want to check of our detected batch is adjusted by performing a hierarchical clustering with that batch.

```{r}
samplevial <- substr(colnames(thca.filt), 14, 16)
table(data.frame(TYPE=thca.filt$type, SV=samplevial))
logCPM <- cpm(dge.filt, log=TRUE, prior.count=3)
d <- as.dist(1-cor(logCPM, method="spearman"))
sampleClustering <- hclust(d)
batch <- as.integer(factor(samplevial))
sampleDendrogram <- as.dendrogram(sampleClustering, hang=0.1)
names(batch) <- colnames(thca.filt)
outcome <- paste(substr(colnames(thca.filt), 9, 12), as.character(thca.filt$type), sep="-")
names(outcome) <- colnames(thca.filt4)
sampleDendrogram <- dendrapply(sampleDendrogram,
                               function(x, batch, labels) {
                                 if (is.leaf(x)) {
                                   attr(x, "nodePar") <- list(lab.col=as.vector(batch[attr(x, "label")]))
                                   attr(x, "label") <- as.vector(labels[attr(x, "label")])
                                 }
                                 x
                               }, batch, outcome)
plot(sampleDendrogram, main="Hierarchical clustering of samples")
legend("topright", paste("Batch", sort(unique(batch)), levels(factor(samplevial))), fill=sort(unique(batch)))
```

We observe that there is still the batch effect caused by samplevial, so we are going to remove this batch effect with an aggresive strategy called *Singular Value Decomposition (SVD)*. 

```{r}
library(corpcor)
s <- fast.svd(t(scale(t(logCPM), center = TRUE, scale = TRUE)))
pcSds <- s$d
pcSds[1] <- 0
svdexp <- s$u %*% diag(pcSds) %*% t(s$v)
colnames(svdexp) <- colnames(thca.filt)
class(svdexp)
dim(svdexp)
```

Visualize the effect of the batch removing with a hiearchical clustering.

```{r sampleClustering, fig.height=7, fig.width=14, dpi=100, echo=TRUE, fig.cap="Figure S12: Hierarchical clustering of the samples by Portion Analyuc."}
d <- as.dist(1 - cor(svdexp, method = "spearman"))
sampleClustering <- hclust(d)
sampleDendrogram <- as.dendrogram(sampleClustering, hang = 0.1)
names(batch) <- colnames(thca.filt)
outcome <- as.character(thca.filt$type)
names(outcome) <- colnames(thca.filt)
sampleDendrogram <- dendrapply(sampleDendrogram, function(x, batch, labels) {
## for every node in the dendrogram if it is a leaf node
if (is.leaf(x)) {
attr(x, "nodePar") <- list(lab.col = as.vector(batch[attr(x, "label")])) ## color by batch
attr(x, "label") <- as.vector(labels[attr(x, "label")]) ## label by outcome
}
x
}, batch, outcome) ## these are the second and third arguments in the function
plot(sampleDendrogram, main = "Hierarchical clustering of samples")
legend("topright", paste("Batch", sort(unique(batch))), fill = sort(unique(batch)))
```

Once, we have adjusted p-values according with batch effect, we want to obtain the truly DE genes

```{r}
tumExp <- rowMeans(logCPM[, thca.filt$type == "tumor"])
norExp <- rowMeans(logCPM[, thca.filt$type == "normal"])
log2fc <- norExp - tumExp
ranking <- order(abs(log2fc), decreasing = TRUE)
genes <- data.frame(LogFC = round(log2fc[ranking], digits = 3), PValue = round(pvsv, digits = 3), row.names = rowData(thca.filt)$symbol[ranking],
check.names = FALSE)
genes$FDRcutoff <- (1:nrow(gen) * 0.05)/nrow(gen)
genes$FDRpval <- p.adjust(gen$PValue, method = "fdr")
maskfdr <- gen$FDRpval < 0.01
maskFC <- gen$LogFC < -2.5 | gen$LogFC > 2.5
genes.fdr <- genes[maskfdr, ]
genes.fc <- genes.fdr[maskFC, ]
genes.fc
```

In this table, we can see all the obtained differential expressed genes (7503) with a adjust p-value with FDR lower than 1%. 

We can plot raw p-values as function on their ranking, and drawing a line through the origin with slope α/m, provides a geometrical explanation for how FDR values are derived and why some of them are identical.

```{r}
alpha <- 0.01 ; plot(1:8, genes.fc$PValue[1:8], pch=19, xlab="Ranking", ylab="Raw p-value", las=1, main = "Plot of raw P-values")
abline(a=0, b=alpha/nrow(genes.fc)) ; abline(v=sum(genes$PValue[1:8]-1:8*alpha/nrow(genes) < 0), lty=2)
```

We also can perform a *Volcano Plot*, widely used diagnostic
plot in DE analysis, to assess the extent of DE by plotting the raw p-values as function of their fold-changes, both in logarithmic scale.

```{r}
plot(genes$LogFC, -log10(genes$PValue), pch=".", cex=3, xlab="Log fold-change", ylab="Raw p-value", las=1)
abline(h=-log10(max(genes$PValue[genes$FDRpvalue <= 0.05])), lty=2)
```

## Session information

```{r, message=FALSE}
sessionInfo()
```