---
output:
  BiocStyle::html_document
---

```{r setup, cache=FALSE, echo=FALSE, results='hide', message=FALSE}
library(knitr)
library(SummarizedExperiment)
library(edgeR)
library(geneplotter)
library(sva)

opts_chunk$set(cache=TRUE,
               cache.path="cache/file2",
               cache.extra=R.version.string,
               autodep=TRUE,
               fig.align="center",
               comment="")

thca <- readRDS("./seTHCA.rds")
dge <- readRDS(file.path("results", "dge.rds"))
thca.filt.unnorm <- readRDS(file.path("results", "thca.filt.unnorm.rds"))
dge.filt.unnorm <- readRDS(file.path("results", "dge.filt.unnorm.rds"))
thca.filt <- readRDS(file.path("results", "thca.filt4.rds"))
dge.filt <- readRDS(file.path("results", "dge.filt4.rds"))
```

# Differential expression

As it is seen in batch identification, there is a cluster of the dataset regarding to the samplevial, which can be confunding with the outcome of interest. Samplevial variable has 5 states with different sample sizes for each one, there is not enough size to adjust the data for that batch. For that reason, a simple examination of expression changes and their associated p-values is going to be performed in order to try to capture sources of heterogeneity, such as non-biological variability, on the filtered and normalized dataset.

First of all, the number of significant differential expressed genes is calculated \@ref(pvalue), this is done without any batch adjustment. 
The FDR approach is used to perform this calculation as some false positives can be accepted in this studie, the aim is to be more laxe.

```{r pvalue}
mod <- model.matrix(~thca.filt$type, data = colData(thca.filt))
mod0 <- model.matrix(~1, data = colData(thca.filt))
pv <- f.pvalue(assays(thca.filt)$logCPM, mod, mod0)
sum(p.adjust(pv, method="fdr") < 0.01)
```

There are `r sum(p.adjust(pv, method="fdr") < 0.01)` genes changing significantly
their expression at FDR < 1%. The distribution of the resulting p-values is shown in Figure \@ref(fig:pdist).

```{r pdist, echo=FALSE, out.width="400px", fig.cap="Supplementary figure 16. Distribution of raw p-values for an F-test on every gene between tumor and normal samples.\\label{fig:plot}"}
hist(pv, main="Histogram before batch adjustment", col = "skyblue", las=1, xlab="p-values", ylab = "Frequency")
```

This histogram shows that most of the genes have a p-value less than 0.1, so they are considered as diferential expressed genes.

The following step is to estimate surrogate variables by the sva method \@rev(sva). 

```{r sva}
assays(thca.filt)$logCPM <- cpm(dge.filt, log=TRUE, prior.count=3)
sv <- sva(assays(thca.filt)$logCPM, mod, mod0)
sv$n
names(sv)
```

`r sv$n` surrogate variables are found. Taking into account this data, the extent of differential expression is assessed adjusting for these surrogate variables \@rev(batchadjust).

```{r batchadjust}
modsv <- cbind(mod, sv$sv)
mod0sv <- cbind(mod0, sv$sv)
pvsv <- f.pvalue(assays(thca.filt)$logCPM, modsv, mod0sv)
sum(p.adjust(pvsv, method="fdr") < 0.01)
```

The number of differential expressed genes has increased to `r sum(p.adjust(pvsv, method="fdr") < 0.01)` with the batch adjustment.
Figure \@ref(fig:psvdist) shows the resulting distribution of p-values. More significant genes with FDR less than 1% can be observed.

```{r psvdist, echo=FALSE, out.width="400px", fig.cap="Supplementary figure 17. Distribution of raw p-values for an F-test on every gene between tumor and normal samples, adjusting for surrogate variables estimated with SVA.\\label{fig:plot}"}
hist(pvsv, main="Histogram after batch adjustment", col = "skyblue", las=1, xlab="p-values", ylab = "Frequency")
```

After the batch adjustement, the previously detected coundufnding effect due to sample vial is checked \@ref(sampleClusteringAdjust).

```{r sampleClusteringAdjust, fig.height=7, fig.width=14, dpi=100, echo=FALSE, fig.cap="Supplementary figure 18. Hierarchical clustering of the samples after adjusting for surrogate variables\\label{fig:plot}"}
samplevial <- substr(colnames(thca.filt), 14, 16)
logCPM <- cpm(dge.filt, log=TRUE, prior.count=3)
d <- as.dist(1-cor(logCPM, method="spearman"))
sampleClustering <- hclust(d)
batch <- as.integer(factor(samplevial))
sampleDendrogram <- as.dendrogram(sampleClustering, hang=0.1)
names(batch) <- colnames(thca.filt)
outcome <- paste(substr(colnames(thca.filt), 9, 12), as.character(thca.filt$type), sep="-")
names(outcome) <- colnames(thca.filt)
sampleDendrogram <- dendrapply(sampleDendrogram,
                               function(x, batch, labels) {
                                 if (is.leaf(x)) {
                                   attr(x, "nodePar") <- list(lab.col=as.vector(batch[attr(x, "label")]))
                                   attr(x, "label") <- as.vector(labels[attr(x, "label")])
                                 }
                                 x
                               }, batch, outcome)
plot(sampleDendrogram, main="Hierarchical clustering of samples")
legend("topright", paste("Batch", sort(unique(batch)), levels(factor(samplevial))), fill=sort(unique(batch)))
```

It can be observed that there is still two clusters due to the batch effect caused by samplevial, sva doesn't work for this variable. For this reason, an agressive strategy called *Singular Value Decomposition (SVD)* is used in order to remove this conunfunding effect \@ref(svd).

Results are again visualized with a Hierarchical Clustering \@ref(sampleClusteringRemov). 

```{r svd}
library(corpcor)
s <- fast.svd(t(scale(t(logCPM), center = TRUE, scale = TRUE)))
pcSds <- s$d
pcSds[1] <- 0
svdexp <- s$u %*% diag(pcSds) %*% t(s$v)
colnames(svdexp) <- colnames(thca.filt)
class(svdexp)
dim(svdexp)
```

```{r sampleClusteringRemov, fig.height=7, fig.width=14, dpi=100, echo=FALSE, fig.cap="Supplementary figure 19: Hierarchical clustering after removing samplevial effect.\\label{fig:plot}"}
d <- as.dist(1 - cor(svdexp, method = "spearman"))
sampleClustering <- hclust(d)
sampleDendrogram <- as.dendrogram(sampleClustering, hang = 0.1)
names(batch) <- colnames(thca.filt)
outcome <- as.character(thca.filt$type)
names(outcome) <- colnames(thca.filt)
sampleDendrogram <- dendrapply(sampleDendrogram, function(x, batch, labels) {
## for every node in the dendrogram if it is a leaf node
if (is.leaf(x)) {
attr(x, "nodePar") <- list(lab.col = as.vector(batch[attr(x, "label")])) ## color by batch
attr(x, "label") <- as.vector(labels[attr(x, "label")]) ## label by outcome
}
x
}, batch, outcome) ## these are the second and third arguments in the function
plot(sampleDendrogram, main = "Hierarchical clustering of samples")
legend("topright", paste("Batch", sort(unique(batch))), fill = sort(unique(batch)))
```

After SVD, a clustering among samples is not observed, removing the batch effect of sample vial is been achieved succesfully. Counfunding results in the outcome of interest are avoided.

Finally, the aim is to obtain the truly differentially expressed genes between tumor and normal samples with significant logFC and p-value.

```{r genes}
tumExp <- rowMeans(logCPM[, thca.filt$type == "tumor"])
norExp <- rowMeans(logCPM[, thca.filt$type == "normal"])
log2fc <- norExp - tumExp
ranking <- order(abs(log2fc), decreasing = TRUE)
genes <- data.frame(LogFC = round(log2fc[ranking], digits = 3), PValue = round(pvsv, digits = 3), row.names = rowData(thca.filt)$symbol[ranking],
check.names = FALSE)
genes$FDRcutoff <- (1:nrow(genes) * 0.05)/nrow(genes)
genes$FDRpval <- p.adjust(genes$PValue, method = "fdr")
head(genes)
```

From the 9423 genes that have passed filters, 7519 genes have a significant p-value (<0.01) in previous analysis, but they are not all differentially expressed genes because some of them have a logFC close to 0 which meand not differnece between groups. 

In order to have a visual idea of the differential expressed genes, a *Volcano Plot* is performed \@ref(VolcanoPlot), a widely used diagnostic
plot in DE analysis, in order to assess the extent of DE by plotting the raw p-values as function of their fold-changes, both in logarithmic scale.

```{r VolcanoPlot, fig.height=5, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Supplementary figure 20. Volcano Plot of normalized genes.\\label{fig:plot}"}
plot(genes$LogFC, -log10(genes$PValue), pch=".", cex=3, xlab="Log 2 fold-change", ylab="-Log 10 p-value", las=1, main="Volcano Plot")
# Add colored points: red if padj<0.05, orange of log2FC>1, green if both)
with(subset(genes, PValue < .05 ), points(LogFC, -log10(PValue),pch=20, col="chocolate1"))
with(subset(genes, abs(LogFC)>2.5), points(LogFC, -log10(PValue), pch=20, col="skyblue1"))
with(subset(genes, PValue<.05 & abs(LogFC)>2.5), points(LogFC, -log10(PValue), pch=20, col="darkolivegreen2"))
abline(h=-log10(max(genes$PValue[genes$FDRpval <= 0.05])), lty=2, col="red")
```

Most of the genes are located vertically to a value of Log2FC close to 0, what means that they are not differential expressed between groups, altough they have a significant p-value (orange points). Points with a high log2FC are also observed, suggesting differential expression but with a non-significant p-value (blue points). Finally, green points are those genes with a high logFC and significant p-value.

As we only are interested in the green genes, there is a need to filter the table by FDR < 0.01, to get only significant genes; and by logFC > abs(2.5) in order to get the truly differentially expressed genes between tumor and normal samples. The table \@ref(DEgenes) shows the 111 DE genes wiht their correspoding values.

```{r DEgenes}
maskfdr <- genes$FDRpval < 0.01
maskfc <- genes$LogFC < -2.5 | genes$LogFC > 2.5
genes.DE <- genes[maskfdr & maskfc, ]
genes.DE
```

A prior bibliographic search if the resulting genes was performed, those genes are verified to be mostly related with thyroid carcinome.

## Session information

```{r, message=FALSE}
sessionInfo()
```