---
title: "Analysis of a TCGA RNA-seq data set on Thyroid Carcinoma"
author:
- name: Lydia Fortea Gonzalez
  affiliation:
  - &id Master Programme on Bioinformatics for Health Sciences, Universitat Pompeu Fabra, Barcelona, Spain
  email: lydia.fortea01@estudiant.upf.edu
- name: Julia Mir Pedrol
  affiliation: *id
  email: julia.mir01@estudiant.upf.edu
- name: Helena Rodriguez Lloveras
  affiliation: *id
  email: helena.rodriguez01@estudiant.upf.edu
date: "`r format(Sys.time(), '%B %e, %Y')`"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
bibliography: bibliography.bib
---

# Introduction

Thyroid cancer (THCA) is the most common endocrine malignancy. It is thought that more than 60% of the
population has thyroid nodules, but only a 8-16% of these will become malignant. Thus,
TC only represents a 2.3% of all cancers worldwide. It affects more women than men. The
Cancer Genome Atlas (TCGA) has comprehensively profiled this type of cancer
in a patient cohort. Here we analyze the expression profiles of those
patients, accessible in the form of a raw RNA-seq counts produced by
@rahman2015alternative using a pipeline based on the R/Bioconductor
software package `r Biocpkg("Rsubread")`.

This document is written in [R markdown](https://rmarkdown.rstudio.com) and
should be processed using R and you need to install the packages
`r CRANpkg("knitr")` and `r CRANpkg("markdown")`. Moreover, it using the official
style for Bioconductor vignettes facilitated by the Bioconductor package
`r Biocpkg("BiocStyle")`. Please consult that package documentation, and particularly
the [vignette](http://bioconductor.org/packages/release/bioc/vignettes/BiocStyle/inst/doc/AuthoringRmdVignettes.html)
on "Authoring R Markdown vignettes", for full details on how to elaborate this
kind of documents.

The specific instructions to generate the final HTML report are written in a
[Makefile](https://en.wikipedia.org/wiki/Makefile). To run it you just need to type

   ```
   $ make
   ```

on the unix shell. The `makefile` contains instructions to separately process
the different files that integrate this report, and so only the file that has
been modified will be rebuilt. If you wish to remove or add files to be processed,
you should modify the `makefile`.

The directory `results` will contain resulting files produced during the analysis,
however, figures are going to end by default directories associated with the
source filenames from where they were created.

```{r echo=FALSE, results='asis'}
cat(readLines("QAanalysis.knit.md"), sep="\n")
```

```{r echo=FALSE, results='asis'}
cat(readLines("DEanalysis.knit.md"), sep="\n")
```

```{r echo=FALSE, results='asis'}
cat(readLines("FEanalysis.knit.md"), sep="\n")
```

# References
